Shallow EDSLs and Object-Oriented Programming

### 2016-06-20 paper/sections/ShallowOO.tex
Section 3 in progress

\begin{grammar}
<circuit> ::= `fan' <positive-number>
\alt `identity' <positive-number>
\alt `beside' <circuit> <circuit>
\end{grammar}

The DSL consists of three constructs - two primitives
\emph{identity} and \emph{fan} and one combinator \emph{beside}.
Their meanings are: \emph{identity n} contains \emph{n} isolated vertical wires;
\emph{fan n} has \emph{n} vertical wires with its first wire connected to
all the remaining wires from top to bottom; $beside\ c_1\ c_2$ joins two circuits
$c_1$ and $c_2$ horizontally.
With these three constructs, simple circuits can be described.
%For example, Figure~\ref{} visualize the circuit \emph{beside (fan 3) (identity 3)}.

@@

\lstinputlisting[linerange=2-2]{./code/shallowCircuit.hs}%APPLY:CIRCUIT_TYPE
\lstinputlisting[linerange=5-7]{./code/shallowCircuit.hs}%APPLY:TYPES
The type \lstinline{Circuit}, representing the semantic domain, is to be filled in with a concrete type according to the semantics.
Suppose that the semantics of \dsl is to calculate the width of a
circuit. The definitions would be:
\lstinputlisting[linerange=11-17]{./code/shallowCircuit.hs}%APPLY:CIRCUIT1
%As shallow embedding is semantics-oriented,
\bruno{why is ``id'' in bold?}
\lstinline{width1} is an identity function because \lstinline{Circuit} is a type
synonym of \lstinline{Int} and it is already an integer after circuit construction.
For example, we will immediately get $6$ for \lstinline{beside (fan 3) (identity 3)}.

@@

A simple rewriting of the previous program is to wrap the result into an
datatype, getting back the value through pattern matching:
\lstinputlisting[linerange=21-28]{./code/shallowCircuit.hs}%APPLY:CIRCUIT2
% Then circuit construction gives back a \lstinline{Circuit2} which contains an
% integer value representing the width and \lstinline{width2} extracts out the value.
% Is this definition still shallow embedding?
%One may be curious whether this variant implementation is still shallow embedding or not. The answer is yes because ... % why it is shallow

@@

\paragraph{Towards OOP}
A simple, \emph{semantics preserving}, rewriting of the above program 
uses a record implemented as a newtype. The record has a sole field representing the result:
\lstinputlisting[linerange=33-38]{./code/shallowCircuit.hs}%APPLY:CIRCUIT3
\bruno{width3 should now be width2} 
%We no longer need a separate \lstinline{width3} function as the field name is the extractor.
The implementation is still shallow since essentially the same program
as the previous one. In fact Haskell \lstinline{newtype} do not add any operational
behaviour to the program, and the two programs are effectivelly the
same.  However, having fields makes the program look more like an 
OOP program.

@@

\begin{figure}
\lstinputlisting[linerange=2-13]{../src/scala/base/Circuit.scala}%APPLY:CIRCUIT_SCALA
\caption{OO implementation of \dsl}
\label{code:base}
\end{figure}

>>>>

\begin{grammar}
<circuit> ::= `fan' <positive-number>
\alt `id' <positive-number>
\alt `beside' <circuit> <circuit>
\end{grammar}

The DSL consists of three constructs - two primitives
\emph{id} and \emph{fan} and one combinator \emph{beside}.
Their meanings are: \emph{id n} contains \emph{n} isolated vertical wires;
\emph{fan n} has \emph{n} vertical wires with its first wire connected to
all the remaining wires from top to bottom; $beside\ c_1\ c_2$ joins two circuits
$c_1$ and $c_2$ horizontally.
With these three constructs, simple circuits can be described.
%For example, Figure~\ref{} visualize the circuit \emph{beside (fan 3) (id 3)}.

@@

\lstinputlisting[linerange=2-2]{./code/shallowCircuit.hs}%APPLY:CIRCUIT_TYPE
\lstinputlisting[linerange=5-7]{./code/shallowCircuit.hs}%APPLY:TYPES
The type \lstinline{Circuit}, representing the semantic domain, is to be filled in with a concrete type according to the semantics.
Suppose that the semantics of \dsl is to calculate the width of a
circuit. The definitions would be:
\lstinputlisting[linerange=11-14]{./code/shallowCircuit.hs}%APPLY:CIRCUIT1
%As shallow embedding is semantics-oriented,
%\bruno{why is ``id'' in bold?}
\lstinline{Circuit} is just a type
synonym of \lstinline{Int} and it is already an integer after circuit construction.
For example, we will immediately get $6$ for \lstinline{beside (fan 3) (id 3)}.

@@

A simple rewriting of the previous program is to wrap the result into an
datatype, getting back the value through pattern matching:
\lstinputlisting[linerange=18-25]{./code/shallowCircuit.hs}%APPLY:CIRCUIT2
% Then circuit construction gives back a \lstinline{Circuit2} which contains an
% integer value representing the width and \lstinline{width2} extracts out the value.
% Is this definition still shallow embedding?
%One may be curious whether this variant implementation is still shallow embedding or not. The answer is yes because ... % why it is shallow

@@

\paragraph{Towards OOP}
A simple, \emph{semantics preserving}, rewriting of the above program 
uses a record implemented as a newtype. The record has a sole field representing the result:
\lstinputlisting[linerange=30-35]{./code/shallowCircuit.hs}%APPLY:CIRCUIT3
%\bruno{width3 should now be width2}
%We no longer need a separate \lstinline{width3} function as the field name is the extractor.
The implementation is still shallow since essentially the same program
as the previous one. In fact Haskell \lstinline{newtype} do not add any operational
behaviour to the program, and the two programs are effectivelly the
same.  However, having fields makes the program look more like an 
OOP program.

@@

\begin{figure}
\lstinputlisting[linerange=2-15]{../src/width/Circuit.scala}%APPLY:CIRCUIT_SCALA
\caption{OO implementation of \dsl}
\label{code:base}
\end{figure}
